AFQ-Insight

### 2019-11-22 reports/manuscript/sections/methods.tex

@@

After scaling and imputation, the tractometry data and target phenotypical data
can be organized in a linear model, $y = \mathbf{X} \cdot \beta$, where $y$ is
the phenotype -- categorical, such as a clinical diagnosis, or numerical, such
as the subject's age. The tractometry data is represented in the feature matrix
$\mathbf{X}$, with rows corresponding to different subjects, and columns
corresponding to diffusion measures at different nodes within each tract. The
relationship between tractometric features and the phenotypic target is
characterized by the coefficients in $\beta$. The feature matrix $\mathbf{X}$
has dimensions $N_s \times (N_t \cdot N_n \cdot N_m)$, where $N_s$ is the number
of subjects, $N_t$ is the number of white matter tracts, $N_n$ is the number of
nodes in each tract, and $N_m$ is the number of diffusion metrics calculated at
each node. Typically, $N_t = 20$, $N_n = 100$, and $2 \le N_m \le 8$, resulting
in $\mathcal{O}(10^4)$ features. Conversely, most dMRI studies have
$\mathcal{O}(10^2)$ participants, yielding a feature matrix that is wide and
short. The high dimensionality of this data requires regularization to avoid
overfitting and generate interpretable results.

In addition to regularizing the coefficients in $\beta$, we also wish to
capitalize on our knowledge of the group structure of the tract profile features
in $\mathbf{X}$. The tract-metric combinations form a natural grouping. For
example, we expect that MD features within the left arcuate fasciculus will be
correlated. Likewise for FA values within the right corticospinal tract (CST)
and so on. This group structure is represented in Fig~\ref{fig:group-structure},
which depicts the linear model $y = \mathbf{X} \cdot \beta$. Thus, we seek a
regularization approach that will fit a linear model with anatomically-grouped
covariates, where we expect to observe sparse coefficients at both the group
level and the within-group level. The sparse group lasso (SGL) is a penalized
regression technique that satisfies exactly these
criteria\cite{simon2013sparse}. It solves for a coefficient vector
$\widehat{\beta}$ that satisfies
\begin{equation}
    \widehat{\beta} = \min_\beta \frac{1}{2}
    \norm*{y - \displaystyle \sum_{\ell = 1}^{m} \mathbf{X}^{(\ell)} \cdot \beta^{(\ell)}}_2^2
    + \lambda_1 \displaystyle \sum_{\ell = 1}^{m} \sqrt{p_\ell} \norm*{\beta^{(\ell)}}_2
    + \lambda_2 \norm*{\beta}_1,
    \label{eq:sgl}
\end{equation}

where $m$ is the number of groups, $\mathbf{X}^{(\ell)}$ is the submatrix of
$\mathbf{X}$ corresponding to group $\ell$, $\beta^{(\ell)}$ is the coefficient
vector for group $\ell$ and $p_\ell$ is the length of $\beta^{(\ell)}$. The
first term is the mean square error loss, $L_{mse}$, as in the standard linear
regression framework. The second and third terms encourage groupwise sparsity
and overall sparsity, respectively. If $\lambda_1 = 0$ and $\lambda_2 = 1$, the
SGL reduces to the traditional lasso\cite{tibshirani1996regression}. Conversely,
if $\lambda_1 = 1$ and $\lambda_2 = 0$, the SGL reduces to the group
lasso\cite{yuan2006model}.

When the phenotypical target variable are categorical, as in the case of
predicting a clinical diagnosis, the SGL is readily adapted to logistic
regression, where the probability of a target variable belonging to an arbitrary
defined ``true'' class is the logistic function of the result of the linear
model,

@@

\begin{figure}[!h]
    \centering
    \includegraphics[width=0.65\textwidth]{dMRI_group_structure.png}
    \caption{{\bf dMRI group structure.}
        The phenotypical target data and tractometric features can be organized
    into a linear model, $y = \mathbf{X} \cdot \beta$. The feature matrix
    $\mathbf{X}$ is color-coded to reveal a natural group structure: the left
    (orange) group contains $k$ features from the inferior fronto-occipital
    fascicle (IFOF), the middle (green) group contains $k$ features from the
    corpus callosum, and the right (blue) group contains $k$ features from the
    uncinate. The coefficients in $\beta$ follow the same natural grouping.
    Fascicle images taken from Ref~\cite{yeatman2012tract}.}
    \label{fig:group-structure}
\end{figure}

\subsection*{Metaparameter optimization}

To select the values of the metaparameters $\lambda$

>>>>

TODO:

@@

After scaling and imputation, the tractometry data and target phenotypical data
can be organized in a linear model:

\begin{equation}
y = \mathbf{X} \cdot \beta,
\label{eq:lm}
\end{equation}

where $y$ is the phenotype -- categorical, such as a clinical diagnosis, or
continuous numerical, such as the subject's age. The tractometry data is
represented in the feature matrix $\mathbf{X}$, with rows corresponding to
different subjects, and columns corresponding to diffusion measures at different
nodes within each tract. The relationship between tractometric features and the
phenotypic target is characterized by the coefficients in $\beta$. In general,
the feature matrix $\mathbf{X}$ has dimensions $S \times (T \cdot N \cdot M)$,
where $S$ is the number of subjects, $T$ is the number of white matter tracts,
$N$ is the number of nodes in each tract, and $M$ is the number of diffusion
metrics calculated at each node. Typically, $T = 20$, $N = 100$, and $2 \le M
\le 8$, resulting in $\sim 4,000 - 16,000$  features. Conversely, many dMRI
studies have between a few dozen and a few hundred subjects, yielding a feature
matrix that is wide and short. Even in cases where more than a thousand subjects
are measured (e.g., in the Human Connectome Project, where 1,200 subjects were
measured \cite{VanEssen2012}), the problem is ill-posed: The high dimensionality
of this data requires regularization to avoid overfitting and generate
interpretable results.

Here, we propose that in addition to regularizing the coefficients in $\beta$,
we can also capitalize on our knowledge of the group structure of the tract
profile features in $\mathbf{X}$. The tract-metric combinations form a natural
grouping. For example, we expect that MD features within the left arcuate
fasciculus to co-vary across individuals. Likewise for FA values within the
right corticospinal tract (CST) and so on. This group structure is represented
in Fig~\ref{fig:group-structure}, which depicts the linear model $y = \mathbf{X}
\cdot \beta$. Thus, we seek a regularization approach that will fit a linear
model with anatomically-grouped covariates, where we expect to observe both
groupwise sparsity: the number of groups (tract/metric combinations) with at
least one non-zero coefficients as well as within-group sparsity: the number of
non-zero coefficients within each non-zero groups. The sparse group lasso (SGL)
is a penalized regression technique that satisfies exactly these
criteria\cite{simon2013sparse}. It solves for a coefficient vector
$\widehat{\beta}$ that satisfies
\begin{equation}
    \widehat{\beta} = \min_\beta \frac{1}{2}
    \norm*{y - \displaystyle \sum_{\ell = 1}^{G} \mathbf{X}^{(\ell)} \cdot \beta^{(\ell)}}_2^2
    + \lambda_1 \displaystyle \sum_{\ell = 1}^{G} \sqrt{p_\ell} \norm*{\beta^{(\ell)}}_2
    + \lambda_2 \norm*{\beta}_1,
    \label{eq:sgl}
\end{equation}

where $G$ is the number of groups $\mathbf{X}^{(\ell)}$ is the submatrix of
$\mathbf{X}$ corresponding to group $\ell$, $\beta^{(\ell)}$ is the coefficient
vector for group $\ell$ and $p_\ell$ is the length of $\beta^{(\ell)}$. In the
tractomtetry setting, $G = T \cdot M$ and $\forall \ell: p_\ell = 100$. The
first term is the mean square error loss, $L_{mse}$, as in the standard linear
regression framework. The second and third terms encourage groupwise sparsity
and overall sparsity, respectively. If $\lambda_1 = 0$ and $\lambda_2 = 1$, the
SGL reduces to the traditional lasso\cite{tibshirani1996regression}. Conversely,
if $\lambda_1 = 1$ and $\lambda_2 = 0$, the SGL reduces to the group
lasso\cite{yuan2006model}.

\subsection*{Implementation, cross-validation and metaparameter optimization}

For given values of $\lambda_1$ and $\lambda_2$, the cost function in equation
\ref{eq:sgl} can be optimized using proximal gradient descent methods
\cite{parikh2014proximal} here implemented as a custom proximal operator that is
then optimized using the COPT library \cite{fabian_pedregosa_2018_1473584}. To
find the values of the metaparameters $\lambda_1$ and $\lambda_2$, we used a
hyperoptimization procedure, implemented using the hyperopt software
\cite{Bergstra_2015}. XXX We need more details here.

To objectively evaluate the model and guard against over-fitting, we used
cross-validation. XXX More here.

\subsubsection*{Incorporating transformations on $y$}

Often, the target variable $y$ is not in the domain in which the linear model
can be best fit to it. Equation \ref{eq:lm} can be slightly modified as:

\begin{equation}
y = f^-1(\mathbf{X} \cdot \beta),
\label{eq:lm}
\end{equation}

where the transformation function $f^-1$ characterizes the transform applied to
the data before fitting the linear coefficients. For example, an often-used
transform is the logarithmic transform:

\begin{equation}
f(y) = log_n(y)
\label{eq:log_nonlinearity}
\end{equation}

In this case, the model is parametrized by one additional fit parameter, $n$.

\subsubsection*{Categorical $y$}
When the phenotypical target variable is categorical, as in the case of
explaining or predicting the presence of a clinical diagnosis, the SGL is
readily adapted to logistic regression, where the probability of a target
variable belonging to an arbitrary defined ``true'' class is the logistic
function of the result of the linear model,

@@

\begin{figure}[!h]
    \centering
    \includegraphics[width=0.65\textwidth]{dMRI_group_structure.png}
    \caption{{\bf dMRI group structure.}
        The phenotypical target data and tractometric features can be organized
    into a linear model, $y = \mathbf{X} \cdot \beta$. The feature matrix
    $\mathbf{X}$ is color-coded to reveal a natural group structure: the left
    (orange) group contains $k$ features from the inferior fronto-occipital
    fascicle (IFOF), the middle (green) group contains $k$ features from the
    corpus callosum, and the right (blue) group contains $k$ features from the
    uncinate. The coefficients in $\beta$ follow the same natural grouping.
    Fascicle image reproduced with permission from Ref~\cite{yeatman2012tract}
    Figure 1.}
    \label{fig:group-structure}
\end{figure}
